```{r setup, echo=FALSE, results='hide'}
opts_chunk$set(message=FALSE, warning=FALSE, background='#F7F7F7', dpi=100, fig.align='center', dev='CairoPNG', tidy=FALSE)
options(width=100, stringsAsFactors=FALSE)
```

# Generating Soil Component Relation Diagrams

Component association is a function of how frequently components co-occur, weighted by component percentage. This summary should generate structures that are related to soil-landscape concepts that were used to construct collections of map units associated with one or more soil survey legends. These diagrams are 2D approximations of the multi-dimensional relationship between components-- where each node is a component, lines (edges) connect components that have been mapped together, colors define groups of components that frequently co-occur, node proximity is proportional to co-occurrence, and size is proportional to "connectivity".

Details pending.

## Simple Example
```{r libraries-and-sample, fig.width=14, fig.height=7, out.width='100%'}
# load libraries
library(soilDB)
library(sharpshootR)
library(plyr)
library(igraph)
library(RColorBrewer)

# load sample data associated with the amador soil series
data(amador)

# map unit keys, component names, component percentages
head(amador)

# convert into adjacency matrix, based on no. times component co-occur
m <- component.adj.matrix(amador, method='occurrence')
print(m)

# convert into adjacency matrix, weighted by component percent
m <- component.adj.matrix(amador)
print(round(m, 2))

# tighten margins
par(mfcol=c(1,2), mar=c(0,0,0,0))
# plot as network diagram, with Amador soil highlighted
plotSoilRelationGraph(m, s='amador')
# plot as dendrogram, with Amador soil highlighted
plotSoilRelationGraph(m, s='amador', type='dendrogram')
```

## NASIS via local database
```{r get-format-nasis-data}
# get legend/DMU/component data from NASIS
# in this case, CA630 data
d <- get_component_data_from_NASIS_db()

# normalize component names
d$compname <- tolower(d$compname)

# remove misc. areas components
d <- subset(d, compkind != 'miscellaneous area')

# remove some higher-taxa and non-soil components
d <- subset(d, ! compname %in% c('lithic haploxeralfs', 'rock outcrop', 'aquic haploxeralfs', 'ultic haploxeralfs', 'riparian'))

# select only those columns that we really need
d <- d[, c('dmudesc', 'compname', 'comppct_r')]

# check top couple of rows
head(d)

# convert into adjacency matrix, weighted by component percentage
m <- component.adj.matrix(d, mu='dmudesc', co='compname', wt='comppct_r')
```


```{r plot-network-graph, fig.width=15, fig.height=15, out.width='100%'}
# tighten margins
par(mar=c(0,0,2,0))
plotSoilRelationGraph(m)
title('CA630 Components (subject to revision)')
```


## SSURGO via SDA
```{r get-format-SSURGO-data, results='hide'}
# select by survey area(s)
q <- "SELECT 
component.mukey, comppct_r, lower(compname) as compname
FROM legend
INNER JOIN mapunit ON mapunit.lkey = legend.lkey
LEFT OUTER JOIN component ON component.mukey = mapunit.mukey
WHERE legend.areasymbol IN ('CA654')
AND compkind IN ('Series', 'Taxadjunct')"

# run query, process results, and return as data.frame object
res <- SDA_query(q)

# convert into adj. matrix using component percentage weights
m <- component.adj.matrix(res)
```

```{r plot-SSURGO-network-graph, fig.width=15, fig.height=15, out.width='100%'}
# tighten margins
par(mar=c(0,0,2,0))
plotSoilRelationGraph(m)
title('CA654 Components')
```

----------------------------
This document is based on `aqp` version `r utils::packageDescription("aqp", field="Version")`, `soilDB` version `r utils::packageDescription("soilDB", field="Version")`, and `sharpshootR` version `r utils::packageDescription("sharpshootR", field="Version")`.