
---
output:
  html_vignette:
    mathjax: null
    jquery: null
    smart: no
---

```{r setup, echo=FALSE, results='hide', warning=FALSE}
library(knitr, quietly = TRUE)
library(printr, quietly = TRUE)
opts_chunk$set(message=FALSE, warning=FALSE, out.width="100%", fig.retina = 2, fig.align='center', dev='png', dev.args=list(pointsize=10, antialias='cleartype'), tidy=FALSE)
options(width=100, stringsAsFactors=FALSE)

knit_hooks$set(htmlcap = function(before, options, envir) {
  if(!before) {
    paste('<p class="caption" style="font-size:85%; font-style: italic; font-weight: bold;">',options$htmlcap,"</p><hr>",sep="")
    }
    })
```


Component Interpretation Summary via SDA
========================================
D.E. Beaudette
<br>
`r format(Sys.time(), "%Y-%m-%d")`
<br>
This document is based on `soilDB` version `r utils::packageDescription("soilDB", field="Version")`.

Lets get some interpretations that are relevant to MLRA 18 (Sierra Nevada Foothills) soils so that we can compare components named [**Auburn**](https://soilseries.sc.egov.usda.gov/OSD_Docs/A/AUBURN.html) and [**Dunstone**](https://soilseries.sc.egov.usda.gov/OSD_Docs/D/DUNSTONE.html), and just for fun lets throw in a very different soil from MLRA 17 (San Joaquin Valley)--[**Hanford**](https://soilseries.sc.egov.usda.gov/OSD_Docs/H/HANFORD.html).

Note that SDA has a 10,000 row limit, so we have to be a little more clever when writing the query.
```{r}
library(soilDB)
library(plyr)
library(reshape2)
library(lattice)

# beware, there are hard limits (10k rows) on what can be returned by SDA
q <- "SELECT component.cokey, compname, mrulename, interplr, interphr 
FROM component INNER JOIN cointerp ON component.cokey = cointerp.cokey
WHERE compname IN ('Auburn', 'Dunstone', 'Hanford')
AND seqnum = 0
AND mrulename IN ('ENG - Construction Materials; Topsoil', 
'ENG - Sewage Lagoons', 'ENG - Septic Tank Absorption Fields', 
'ENG - Unpaved Local Roads and Streets', 
'AGR - California Revised Storie Index (CA)', 
'AGR - Pesticide Loss Potential-Leaching')
AND interplr IS NOT NULL;"

# query and check
x <- SDA_query(q)
head(x)
```

OK, so how to the fuzzy numbers (`interplr` values) for these components compare? Box and whisker plots are the simplest, but sometimes density plots are helpful for viewing the complete distribution. Looks like  
```{r, fig.width=8, fig.height=10}
# compute number of rules and soils
n.soils <- length(unique(x$compname))
n.rules <- length(unique(x$mrulename))

# compare population with box-whisker plot
bwplot(compname ~ interplr | mrulename, data=x, layout=c(1,n.rules))

densityplot(~ interplr | mrulename, groups=compname, data=x, layout=c(1,n.rules), auto.key=list(points=FALSE, lines=TRUE, columns=n.soils), scales=list(y=list(relation='free')))
```


Those figures were useful, sometimes it is nice to see the median values for rule (rows) and component (columns).
```{r}
s <- ddply(x, c('mrulename', 'compname'), summarize, 
           low=quantile(interplr, probs=0.05, na.rm=TRUE), 
           rv=quantile(interplr, probs=0.5, na.rm=TRUE), 
           high=quantile(interplr, probs=0.95, na.rm=TRUE))

kable(dcast(s, mrulename ~ compname, value.var = 'rv'), caption = "Median Fuzzy Ratings")
```


To be continued.
