<title>Soil Database Interface: the soilDB package</title>
# Soil Database Interface: the <tt>soilDB</tt> package

```{r setup, echo=FALSE, results='hide'}
# setup
opts_chunk$set(message=FALSE, warning=FALSE, background='#F7F7F7', dpi=100, fig.align='center', dev='CairoPNG')
options(width=100, stringsAsFactors=FALSE)
```


## Introduction
This package provides methods for extracting soils information from local PedonPC and AK Site databases (MS Access format), local NASIS databases (MS SQL Server), and the SDA webservice. Currently USDA-NCSS data sources are supported, however, there are plans to develop interfaces to outside systems such as the Global Soil Mapping project.
	
It can be difficult to locate all of the dependencies required for sending/processing SOAP requests, especially on UNIX-like operating systems. Windows binary packages for the dependencies can be found [here](http://www.stats.ox.ac.uk/pub/RWin/bin/windows/contrib/2.14/).

This document is based on `aqp` version `r utils::packageDescription("aqp", field="Version")` and `soilDB` version `r utils::packageDescription("soilDB", field="Version")`.


### Inteded Users


### R Notes
A basic understanding of R syntax, object structure, package management system, and programming style will greatly help with integration of the `soilDB` and `aqp` packages into standard workflows. Here are some reminders. Recall that online help documents can be accessed with the syntax `?mean`, which would load the manual page for the "mean()" function. Two `??` can be used to perform fuzzy searching for functions that contain matching keywords.


```{r R-notes, eval=TRUE}
# 1. objects: creation, structure, etc.
x <- 1:10       # 'x' is now a vector of integers 1 to 10 
y <- rnorm(10, mean=0, sd=1)  # 'y' is now a vector of 10 random numbers {mean=0, sd=1}
d <- data.frame(x, y) # d is a rectangular table with columns 'x' and 'y'

str(d) # check structure
class(d) # inspect object class
head(d) # view first 6 rows
rm(d, x, y) # clean-up by deleting these objects

# 2. most function are vectorized: i.e. automatic iteration
x <- 1:10
x + 1 # the '+' function knows about vectors, and recycles scalars accordingly

# 3. missing data can cause problems unless accounted for
x <- c(x, NA) # 'c()' is the concatonate function
# print the value by typing it into the R console followed by enter
x

mean(x) # result is NA, as most functions return NA in the presence of missing data
mean(x, na.rm=TRUE) # mean computed after removal of missing data
```


### Design

The `soilDB` package provides two layers of functionality for extracting data from file-based databases (PedonPC), relational databases (local NASIS), and online data sources (Soil Data Access). High-level functions `fetchPedonPC()` and `fetchNASIS()` query, combine, verify horizon logic, and return `SoilProfileCollection` class objects containing the most commonly used site/pedon/horizon data. Low-level functions such as `get_site_data_from_pedon_db()` and `get_hz_data_from_pedon_db()` extract specific data from a designated source, and return the results as a `data.frame` class object.

### High-Level Functions

**Fetch Pedons from Local NASIS**
The `fetchNASIS()` function will load all pedons from the local NASIS database. Pedon records are returned for *all* pedons in the local database, regardless of the selected set. An ODBC connection named 'nasis_local' must **first** be established (details pending).

```{r local-nasis-demo, eval=FALSE}
library(soilDB)
# this will only work on Windows after you have setup a ODBC connection
# to the local NASIS database (SQL Server 2008), and named it 'nasis_local'
s <- fetchNASIS()

# subset 'Gopheridge' pedons
subset.idx <- grep('gopheridge', gopheridge$correlated_as, ignore.case=TRUE)
gopheridge <- s[subset.idx, ]

# plot the data
plot(gopheridge, name='hzname')
```


**Fetch Pedons from PedonPC 5.x Database**
The `fetchPedonPC(dsn)` function will load all pedon records into the current R session, based on the value of `dsn`-- a path to a PedonPC version 5.x database.

```{r pedonPC-demo, eval=FALSE}
library(soilDB)
# this will only work if you have a PedonPC Database and Windows :(
# get all of the data from a PedonPC 5.x database (you will have to adjust the DSN)
dsn <- "S:/Service_Center/NRCS/pedon/pedon.accdb"
s <- fetchPedonPC(dsn)

# subset 'Gopheridge' pedons
subset.idx <- grep('gopheridge', gopheridge$correlated_as, ignore.case=TRUE)
gopheridge <- s[subset.idx, ]

# plot the data
plot(gopheridge, name='hzname')
```


### Low-Level Functions

**Fetch Data from a PedonPC 5.x Database**
These functions can be used to extract only subsets of the data available in a PedonPC database, but are generally only useful for advanced users.

```{r pedonPC-low-level-functions, eval=FALSE}
# fetch only site data
site_data <- get_site_data_from_pedon_db(dsn)
# fetch only horizon data
hz_data <- get_hz_data_from_pedon_db(dsn)
# fetch only color data
color_data <- get_colors_from_pedon_db(dsn)
# fetch extended data:
# diagnostic horizons, fragment summary, texture modifiers, geomorphology, taxonomic history
extended_data <- get_extended_data_from_pedon_db(dsn)
```

**Fetch Data from a Local NASIS Database**
These functions can be used to extract only subsets of the data available in the local NASIS database, but are generally only useful for advanced users.

```{r local-nasis-low-level-functions, eval=FALSE}
# fetch only site data
site_data <- get_site_data_from_NASIS_db()
# fetch only horizon data
hz_data <- get_hz_data_from_NASIS_db()
# fetch only color data
color_data <- get_colors_from_NASIS_db()
# fetch extended data:
# diagnostic horizons, fragment summary, texture modifiers, geomorphology, taxonomic history
extended_data <- get_extended_data_from_NASIS_db()
```

## Examples

### Gopheridge Pedons

Open R, and setup the environment by loading packages and the sample dataset. Missing libraries can be installed via `install.packages('soilDB', dep=TRUE)`.
```{r gopheridge-demo-1, eval=TRUE}
# libraries: you may have to install these:
library(soilDB)
library(Hmisc)
library(ape)
library(lattice)
library(cluster)
library(reshape)

# load example dataset
data(gopheridge)

# what kind of object is this?
class(gopheridge)

# print method of SoilProfileCollection objects
# lots in there!
gopheridge
```


Generate some aggregates of horizon data, and save as site-level data.
```{r gopheridge-demo-2, eval=TRUE}
# compute soil depth using an in-line function, note that max(SPC) returns the bottom-most depth
# this may not be the same as 'soil depth' when Cr or R horizons are present
gopheridge$soil.depth <- profileApply(gopheridge, function(x) max(x))

# horizon-thickness weighted mean (beware of missing data!)
f.wt.prop <- function(x, prop) {h <- horizons(x); w <- with(h, hzdepb-hzdept); wtd.mean(h[[prop]], weights=w)}

# get hz mid-point of hz with max clay content
f.max.clay.depth <- function(x) {
	h <- horizons(x) 
	max.clay <- which.max(h$clay) 
	with(h[max.clay, ], (hzdept+hzdepb) / 2)
	}

# apply by-profile, returning a vector, assigned to new column in @site
gopheridge$wt.clay <- profileApply(gopheridge, f.wt.prop, prop='clay')
gopheridge$wt.rf <- profileApply(gopheridge, f.wt.prop, prop='total_frags_pct')
gopheridge$max.clay.depth <- profileApply(gopheridge, f.max.clay.depth)
```

Define functions for weighted-mean properties within the PCS or diagnostic horizon, then apply to each profile. Also extract diagnostic horizon thickness and top depths, then save as site-level data.
```{r gopheridge-demo-3, eval=TRUE}
# compute aggregate (wt.mean) clay within particle-size control section
# note that this requires conditional eval of data that may contain NA
# see ?slab and ?soil.slot for details on the syntax
f.pcs.prop <- function(x, prop) {
  # these are accessed from @site
  sv <- c(x$psctopdepth, x$pscbotdepth)
  # test for missing PCS data
  if(any(is.na(sv)))
    return(NA)

  # create formula from named property
  fm <- as.formula(paste('~', prop))
  # return just the (weighted) mean, accessed from @horizons
  s <- slab(x, fm, seg_vect=sv)$p.mean
  return(s)
}

# compute the weighted-mean of some property within a given diagnostic horizon
# note that this requires conditional eval of data that may contain NA
# see ?slab and ?soil.slot for details on the syntax
# note that function expects certain columns within 'x'
f.diag.wt.prop <- function(x, d.hz, prop) {
	# extract diagnostic horizon data
	d <- diagnostic_hz(x)
	# subset to the requested diagnostic hz
	d <- d[d$diag_kind == d.hz, ]
	# if missing return NA
	if(nrow(d) == 0)
		return(NA)
	
	# extract depths and check for missing
	sv <- c(d$featdept, d$featdepb)
	if(any(is.na(sv)))
		return(NA)
	
	# create formula from named property
	fm <- as.formula(paste('~', prop))
	# return just the (weighted) mean, accessed from @horizons
	s <- slab(x, fm, seg_vect=sv)$p.mean
	return(s)
}

# conditional eval of thickness of some diagnostic feature or horizon
# will return a vector of length(x), you can save to @site
f.diag.thickness <- function(x, d.hz) {
	# extract diagnostic horizon data
	d <- diagnostic_hz(x)
	# subset to the requested diagnostic hz
	d <- d[d$diag_kind == d.hz, ]
	# if missing return NA
	if(nrow(d) == 0)
		return(NA)
	
	# compute thickness
	thick <- d$featdepb - d$featdept
	return(thick)
}

# conditional eval of top hz depth of diagnostic feature or horizon
# will return a vector of length(x), you can save to @site
f.diag.top <- function(x, d.hz) {
	# extract diagnostic horizon data
	d <- diagnostic_hz(x)
	# subset to the requested diagnostic hz
	d <- d[d$diag_kind == d.hz, ]
	# if missing return NA
	if(nrow(d) == 0)
		return(NA)
	
	# return top depth
	return(d$featdept)
}


# compute wt.mean clay within PCS, save to @site
gopheridge$pcs.clay <- profileApply(gopheridge, f.pcs.prop, prop='clay')

# compute wt.mean clay within argillic horizon, save to @site
gopheridge$argillic.clay <- profileApply(gopheridge, f.diag.wt.prop, d.hz='argillic horizon', prop='clay')

# compute argillic hz thickness, save to @site
gopheridge$argillic.thick <- profileApply(gopheridge, f.diag.thickness, d.hz='argillic horizon')

# compute depth to paralithic contact, if present, save to @site
gopheridge$paralithic.contact.depth <- profileApply(gopheridge, f.diag.top, d.hz='paralithic contact')
```



Re-order the `gopheridge` profiles by presence/absense of paralithic contact, plot, and then annotate depth of Cr contact.
```{r basic_plotting, fig.width=10, fig.height=5, out.width='100%'}
# order by presence of paralithic contact
new.order <- order(gopheridge$paralithic.contact)
# setup margins:
par(mar=c(0,0,0,0))
plot(gopheridge, name='hzname', plot.order=new.order)
# annotate paralithic contact with lines
x.pos <- 1:length(gopheridge)
lines(x.pos, gopheridge$paralithic.contact.depth[new.order], lty=2, col='black')
```
